Computing likelihoods under the SMC for a general class of ARGs.

### 2023-12-13 paper.tex
update intro

\newcommand{\tsinfer}[0]{\texttt{tsinfer}}
\newcommand{\kwarg}[0]{\texttt{KwARG}}
\newcommand{\argweaver}[0]{\texttt{ARGweaver}}
\newcommand{\argweaverD}[0]{\texttt{ARGweaver-D}}
\newcommand{\relate}[0]{\texttt{Relate}}
\newcommand{\espalier}[0]{\texttt{Espalier}}
\newcommand{\arbores}[0]{\texttt{Arbores}}
\newcommand{\tskit}[0]{\texttt{tskit}}

@@

[TL;DR] Importance of ARGs.
There is a discrepancy between ARGs as object of theoretical reasoning 
vs what can be known. Explain the difference between precision/detail and accuracy. 
High precision can give a false sense of certainty in the provided estimate.

In population genetics, the questions we try to answer are generally 
about the unobserved processes that gave rise to the genetic variation 
we can observe given a set of samples. 
The coalescent with recombination 
Kingman, 1982a,b; Tajima, 1983; Hudson, 1983) is generally used as 
the generative model that allows us to understand the genealogy 
underlying a particular sampleset. 
Datastructure that captures it: ARG.

For a very long time people have assumed that what is meant by an ARG is the 
fully specified genealogical history where all events are uniquely positioned 
both in time and along the genome. Yet the ease by which the CwR can be simulated,
and therefore the ease by which an ARG can be generated, is not not matched by 
the usability of the model for inference.

@@

Both approaches have further formed the basis (of many) of the recent advancements in 
scalable ARG inference methods [references]. A real distinction can be made however 
between \argweaver, the first real breakthrough in scalable ARG inference, and the 
more recent methods. The ARGs sampled by \argweaver from the posterior distribution 
given the observed variational data, are all fully precise and detailed in the sense 
that all events are explicitly inferred. Note however that the set of possible node times 
has to be limited to enable the usage of an HMM. This fully detailed inference holds 
for recombination, but also for migration events \cite{hubisz_2020_mapping}. 
All other methods, on the other hand, derive their scalability at least to a 
certain degree from the fact that 
the ARGs they infer are approximate structures. In particular, recombination is an 
emergent property of the inferred ARGs explaining tree topology changes going 
from one local tree to the next. The recombination events themselves are not 
explicitly inferred and are thus not part of the resulting data structure.

\subsection{Recent progress in ARGs through approximate structures}
% onto other features

The computational gains from not having to pinpoint recombination events tie in 
with the inevitable uncertainty associated with ARG reconstruction [Hayman, 2022].

Always events that cannot be known. Different ways of dealing with this.
Some approaches have embraced the approximate nature of the resulting ARG to for 
example highlight the inability to resolve certain splits (polytomies).

others rigorous treatment of uncertainty (Arbores, ARGinfer): fully precise about recombination.

% onto ARG cannot be fully known
Note that approximate structure need not be an issue, the fully precise and detailed 
ARG can never be known. Instead, the main question is how to connect these 
approximate structures to existing models.
While this is not a problem when describing the results of simulations 
(where all details are perfectly known), it is a major issue when we wish to formally 
describe the output of inference methods, particularly as datasets approach the population 
scale. 
As discussed in the introduction, the precise details of all 
events in these vast ARGs can never be known, and a data structure that enforces 
complete precision is therefore an impediment to progress.

@@

These methods
have raised the realistic prospect of ARG-based analysis becoming a standard 
part of the population and statistical genetics toolkit.

These recent developments have also meant however that
we are now using the term ARG for a varying range of approximate structures 
as generated by these different inference methods \citep{wong_a-general_2023}.

Besides not having a unified output format, the degree of completeness 
in which genetic inheritance from ancestors to 
descendants is documented by each of these methods varies extensively.
This has complicated the ability to connect the output of this diverse range of 
ARG inference methods to a model-based deescription that would allow us to perform  
probabilistic exploration of the ARG space around the point estimate, 
or likelihood-based inference (that does not just consider individual marginal trees).
% Existing formalisms assume that an ARG is known in full detail

@@

\begin{figure}[!ht]
\centering
\includegraphics[width=\textwidth]{figures/3-slices.png}
\caption{Simulated ARG under Hudson coalescent (left). $N_e$ varied was constant within each of the 3 time slices: $[(0:1000): 10 000, (1000:5000): 2500, (5000:\infty):10 000]$. Remaining parameters: L=10kb, 20 diploid samples, r=2.5e-8, 148 edges in ARG). A single $N_e$ value was estimated for each slice of the ARG as defined by the true population size change time points. The inferred posterior distribution (right) using the SMC-likelihood was obtained using a uniform prior.}
 \label{fig:3-arg-slices}
\end{figure}

@@

Here, we have introduced a scalable algorithm to compute likelihoods 
under the SMC for a general class of ARGs. By formulating the SMC backwards in time 
we were able to associate a single likelihood value with each edge in the ARG. 
The compatibility of the algorithm with the \texttt{tskit} library allows us to 
exploit its associated efficiencies both for the current work and its future 
applications. This work will contribute significantly 
to ARG-based analysis becoming a standard part of the population 
and statistical genetics toolkit. 
In particular, we see three main applications for the work presented here. 
Firstly, the ability to associate inferred ARGs with a likelihood given a demographic 
model and an associated set of parameters will help bridge the gap between the tremendous recent 
advances in ARG inference and the current state of the art statistical inference of 
past demography. 
Secondly, inferred ARGs can be used as a good starting point for a future MCMC sampler.
Thirdly, hybrid ARG-inference methods that combine the best of both 
heuristic and model-based inference approaches are now possible.
Finally, using a model we can quantify the uncertainty inherently associated 
with ARG-reconstruction.

@@

% 2/ great starting point for MCMC
Any inferred ARG can be used a good starting point 
for a future MCMC-sampler. This potentially implies a massive runtime reduction by 
reducing the necessary burn-in period. Furthermore, by integrating out the exact time 
to the recombination event we further restrict the size of the ARG space that needs to 
be explored. This MCMC-sampler would again rely on the tree sequence data structure to 
efficiently generate and evaluate any new proposal (see \citep{mahmoudi_bayesian_2022}).
[TO DO:]

% 3/ hybrid ARG-inference: scalability + bringing ARG down from pedigree dominated phase
Hybrid backwards-in-time simulations that combine Wright-Fisher dynamics in the 
very recent past and coalescent simulations for the remaining time are a tried and tested 
approach \citep{bhaskar_distortion_2014, nelson_accounting_2020} to enable large scale 
simulations while avoiding the documented biases of the coalescent relative to the 
Wright-Fisher model \citep{bhaskar_distortion_2014, wakeley_gene_2012}. The same approach 
could be used for ARG-inference where one can rely on a heuristic inference method for 
the recent past, while using a model-based approach beyond that cutoff point.\\

% 4/ quantifying uncertainty [unfinished]:
Both of these future directions tie in with the desired ability of any ARG-related method to 
quantify the uncertainty associated with ARG inference

This will in particular enable us to deal with those 
parts in both time and along the genome for which we have insufficient information to confidently 
reconstruct the ARG. The presence of polytomies in ARGs inferred by \tsinfer, for example, 
represents uncertainty. Systematicallty breaking them by sampling from the coalescent rather 
then randomly would already be a major step forward.

Quantifying uncertainty would not necessarily always have to be associated 
with sampling 
with dealing with the entire subset of all compatible genealogies. Uncertainty could also 
be quantified and passed on as metadata to a parent node representing the uncertainty 
in the interval spanned by the polytomy.
% can something similar be done for stacked recombinations?

[TO DO] Comment on accuracy. The described algorithm has been implemented in such a way that 
any valid tree sequence will result in a likelihood value. This also holds for any tree sequence 
that does not meet the assumptions. We see it as future work to determine to what extent 
particular deviations from these assumptions will affect the shape of the likelihood surface given 
a set of parameters.

% final conclusion

>>>>

\newcommand{\tsinfer}[0]{\texttt{tsinfer}}
\newcommand{\kwarg}[0]{\texttt{KwARG}}
\newcommand{\argneelde}[0]{\texttt{ARG-Needle}}
\newcommand{\argweaver}[0]{\texttt{ARGweaver}}
\newcommand{\argweaverD}[0]{\texttt{ARGweaver-D}}
\newcommand{\relate}[0]{\texttt{Relate}}
\newcommand{\espalier}[0]{\texttt{Espalier}}
\newcommand{\arbores}[0]{\texttt{Arbores}}
\newcommand{\tskit}[0]{\texttt{tskit}}

@@

Knowledge of the exact genealogy of a set of samples would make certain problems 
in population genetics (e.g.\ inferring recombination rates) trivial. 
Multitude of methods have started to 
demonstrate the value of attempts to capture the details of these intricately 
interwoven paths of inheritance referred to as an ancestral recombination graph (ARG).
Recent advancements: the inference of ARGs at population scale.
Yet even though potential of ARGs in population and statistical genetics has been 
demonstrated it remains unclear how to connect these reconstructed ARGs to 
the stochastic/generative models that have lead us to study these datastructures 
in the first place.

[Message needs to be here: way to think about genealogies is tied to the CwR. However,
CwR does not enable us to simply compute a likelihood in the absence of a fully 
detailed history. The way forward instead has always been driven by approximations.]

% The coalescent with recombination 
% Kingman, 1982a,b; Tajima, 1983; Hudson, 1983) is generally used as 
% the generative model that allows us to understand the genealogy 
% underlying a particular sampleset. 
% Datastructure that captures it: ARG.

For a very long time people have assumed that what is meant by an ARG is the 
fully specified genealogical history where all events are uniquely positioned 
both in time and along the genome [Wong et al. 2023].
Yet, the ease by which such a fully detailed history can be simulated under the 
CwR, is not matched by the usability of the model for inference. 
In particular, given an ARG, there will always be an infinite amount of other 
ARGs that could have generated the observed genomic dataset 
[potentially with very different likelihoods under the CwR]
\citep{mcvean_2005_approximating}.

% There is a discrepancy between ARGs as object of theoretical reasoning 
% vs what can be known. Explain the difference between precision/detail and accuracy. 
% High precision can give a false sense of certainty in the provided estimate.

% In population genetics, the questions we try to answer are generally 
% about the unobserved processes that gave rise to the genetic variation 
% we can observe given a set of samples.

@@

Both approaches have further formed the basis (of many) of the recent advancements in 
ARG inference methods [Rasmussen et al. 2014, Heine et al. 2018, 
Kelleher et al. 2019, Speidel et al. 2019,
Rasmussen et al. 2022, Zhang et al.2023]. 
A further classification of these methods could be made based on the level of detail 
that is provided on recombination events. The first real breakthrough in large-scale 
ARG inference, \argweaver, heads the first group. 
The ARGs sampled by \argweaver from the posterior distribution 
given the observed variational data, are all fully precise and detailed in the sense 
that all events are explicitly inferred. Note however that the set of possible node times 
has to be limited to enable the usage of an HMM. The other more recent methods, although 
still SMC-based, have added heuristic pre- and/or post-processing steps 
that have allowed them to extend their approach to larger sample sizes. We should also 
mention \kwarg and \arginfer in this category of methods producing fully detailed ARGs 
although these are not based on the SMC or the copying algorithm. In fact they represent 
two extremes along the approximation continuum with \kwarg being entirely 
parsimony-based while \arginfer is a CwR-based MCMC sampler.

The second group of methods (\tsinfer, \relate, \argneedle) has marked a significant  
leap in scalability. These methods derive their scalability at least to a certain 
degree from the fact that the ARGs they infer are approximate structures. In particular, 
recombination is an emergent property of the inferred ARGs explaining tree topology 
changes going from one local tree to the next. The recombination events themselves 
are not explicitly inferred and are thus not part of the resulting data structure.

\subsection{Recent progress in ARGs through approximate structures}

The computational gains from not having to pinpoint recombination events tie in 
with the inevitable uncertainty associated with ARG reconstruction [Hayman, 2022, ADD REFS].

Downsides:
Point estimate:
Some approaches have embraced the approximate nature of the resulting ARG by, for 
example, highlighting the inability to resolve certain splits and instead representing 
them as polytomies \citep{kelleher_2019_inferring}. 
+
This means that the degree of 
completeness in which genetic inheritance from ancestors to descendants is documented 
by each of these methods varies extensively.

But: Advantage of being able to infer ARGs at scale.

Connect this to likelihood approach.

While this is not a problem when describing the results of simulations 
(where all details are perfectly known), it is a major issue when we wish to formally 
describe the output of inference methods, particularly as datasets approach the population 
scale. 
Only successful if we can connect these to statistical models.

Progress driven by approximation: here similarly

Note that approximate structure need not be an issue, the fully precise and detailed 
ARG can never be known. Instead, the main question is how to connect these 
approximate structures to existing models.

@@

This has complicated the ability to connect the output of this diverse range of 
ARG inference methods to a model-based deescription that would allow us to perform  
probabilistic exploration of the ARG space around the point estimate, 
or likelihood-based inference.
The problem is that there is no direct way of calculating the coalescent 
likelihood from a set of marginal genealogies (i.e. integrating over the 
possible ARGs that could have given the set of marginal genealogies).

@@

\begin{figure}[!ht]
\centering
\includegraphics[width=\textwidth]{figures/3-slices.png}
\caption{Simulated ARG under Hudson coalescent (left). $N_e$ varied was constant within each of the 3 time slices: $[(0:1000): 10 000, (1000:5000): 5000, (5000:\infty):10 000]$. Remaining parameters: L=10kb, 20 diploid samples, r=2.5e-8, 148 edges in ARG). A single $N_e$ value was estimated for each slice of the ARG as defined by the true population size change time points. The inferred posterior distribution (right) using the SMC-likelihood was obtained using a uniform prior. Number of edges per time slice: [68, 57, 23]. Locally unary nodes were omitted for clarity.}
\label{fig:3-arg-slices}
\end{figure}

@@

Here, we have introduced a scalable algorithm to compute likelihoods 
under the SMC for a general class of ARGs. By formulating the SMC backwards in time 
we were able to associate a single likelihood value with each edge in the ARG. 
The compatibility of the algorithm with the \texttt{tskit} library allows us to 
exploit its associated efficiencies both for the current work and its future 
applications. This work will contribute significantly 
to ARG-based analysis becoming a standard part of the population 
and statistical genetics toolkit. 
In particular, we see three main applications for the work presented here. 
Firstly, the ability to associate inferred ARGs with a likelihood given a demographic 
model and an associated set of parameters, will help bridge the gap between the tremendous recent 
advances in ARG inference and the current state of the art statistical inference of 
past demography. 
Secondly, hybrid ARG-inference methods that combine the best of both 
Thirdly, inferred ARGs can be used as a good starting point for a future MCMC sampler.
heuristic and model-based inference approaches are now possible.
Finally, using a model we can quantify the uncertainty inherently associated 
with ARG-reconstruction.

@@

% 2/ hybrid ARG-inference: scalability + bringing ARG down from pedigree dominated phase
Hybrid backwards-in-time simulations that combine Wright-Fisher dynamics in the 
very recent past and coalescent simulations for the remaining time are a tried and tested 
approach \citep{bhaskar_distortion_2014, nelson_accounting_2020} to enable large scale 
simulations while avoiding the documented biases of the coalescent relative to the 
Wright-Fisher model \citep{bhaskar_distortion_2014, wakeley_gene_2012}. The same approach 
could be used for ARG-inference where one can rely on a heuristic inference method for 
the recent past, while using a model-based approach beyond that cutoff point.\\

% 3/ great starting point for MCMC
Any inferred ARG can be used a good starting point 
for a future MCMC-sampler. This potentially implies a massive runtime reduction by 
reducing the necessary burn-in period. Furthermore, by integrating out the exact time 
to the recombination event we further restrict the size of the ARG space that needs to 
be explored. This MCMC-sampler would again rely on the tree sequence data structure to 
efficiently generate and evaluate any new proposal (see \citep{mahmoudi_bayesian_2022}).

% 4/ quantifying uncertainty [unfinished]:
Acknowledging the uncertainty associated with ARG inference does not 
necessarily require a full-blown MCMC sampler. Instead, recognising those parts of 
the ARG for which we lack information for confident reconstruction and resolving 
them stochastically by sampling from the coalescent rather than completely arbitrarily 
would already be a major step forward.
Furthermore, quantifying uncertainty would not necessarily always have to be associated 
with sampling from subset of all compatible genealogies. Uncertainty could also 
be quantified and passed on as metadata to the nodes in the data format storing the ARG.
% can something similar be done for stacked recombinations?

As a final remark, the described algorithm relies heavily on the presence of locally unary 
nodes in the tree sequence to correctly identify the full extent of all lineages going back 
through time. Although their presence is crucial to our ability embed any local tree in the 
ARG, the concept and their importance are still relatively new (but see 
\citet{Wong-2023}). Although we have briefly touched upon the observation that in 
the presence of unary nodes we seem to augment the left-to-right logic of the SMC 
with some additional information. 
We believe that the topic of locally unary nodes in general requires further exploration. 
And more specifically, we see it as future work to fully quantify to what extent the 
absence of these locally unary nodes will affect the shape of the likelihood surface given 
any set of parameters.\\

[ TO DO final conclusion]
